---
title: "Getting Started with NestNetR"
author: "Author1, Author2"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NestNetR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
input_file <- knitr::current_input(dir = TRUE)
if (!is.null(input_file)) {
  knitr::opts_knit$set(root.dir = normalizePath(file.path(dirname(input_file), "..")))
}
```
# Getting started {#start}
To classify breeding behaviour from light-level geolocator data in R we need a couple of R packages as well as functions that allow to run our code. 

The package can be installed from GitHub with _pak_.

```{r, eval = FALSE}
install.packages("pak")
pak::pak("bennett-stolze/NestNetR")
```

You should now be able to load the package. We recommend to check (every now and then), if there is a new version of _NestNetR_ available. And if that is the case, re-install the package using the same code you used for initial installation.

```{r, message=FALSE, warning=FALSE}
library(NestNetR)
``` 


# 1 - Loading data
## 1.1 - Raw data
The first step is to load your raw data into R. Different geolocator types (e.g. from different manufacturers or different series) provide raw data in different formats. And while there are functions available to read a whole range of formats, you may have to either write your own function, use simple read text utilities or get in touch with the package managers to write code that fits your format if it is not yet implemented.

The most frequently used geolocators provide files with the extension `.lux` (Migrate Technology Ltd), `.lig` (BAS, Biotrack) or `.glf` (Swiss Ornithological Institute). The functions `read_light` allows you to automatically read these different type of files. In most cases the raw data is stored in a text file that can also be read in to R using the base function `read.table()`.

With the suggested data structure, we can then define metadata information on the individual and the species and define the sub-folder for saving and extracting data files.

```{r}
ID <- "BB959"
Species <- "RuddyTurnstone" 
wd <- "data"
dir.raw <- file.path(wd, "RawData", Species)
```

By using the above metadata we can use the `file.path` and `paste0` commands to include this information in reading and writing of files.
The columns of our data need to be in a specific format with Date being a `POSIXct` class and Light, Tmin & Tmax (and potentially other variables) being numeric values. Check if the structure of your data follows the required format with the function `str`. If not, adjust Date format with `as.POSIXct(raw_light$Date, tz = "GMT")`. We will load both light and temperature data to check their structure.

```{r}
raw_light <- read_light(file.path(wd, "RawData", Species, paste0(ID, ".lux")))
head(raw_light)
str(raw_light)
```

```{r}
raw_deg <- read_deg(file.path(wd, "RawData", Species, paste0(ID, ".deg")))
head(raw_deg)
str(raw_deg)
```

If a `.deg` file matching the full ID is not found, `read_deg()` also tries the base ID with bracketed suffixes removed, for example `BB880(1)` -> `BB880`.

## 1.2 - Identify breeding period
Since the raw data covers the entire migration route, the dataset must be trimmed to the breeding period before processing. This can be done either by automatically detecting the period using a corresponding twilight file (.twl) or by manually specifying the range.

### 1.2.1 - Automatic detection
If a twilight file is available, the breeding period can be automatically detected using the `auto = TRUE` setting in the `set_breeding_period` function. Twilight files (`.csv`) can be created by functions associated with the _GeoLight_ or _TwGeos_ packages. The twilight file needs to be put into the `RawData` folder alongside the light and temperature data.

*Automatic Approach*
```{r}
tm.breeding <- set_breeding_period(dir.raw, raw_light, raw_deg, ID, auto = TRUE)
tm.breeding
```

If no twilight file is available, the breeding period can be manually defined using the argument `auto = FALSE`. In this case, an interactive plotting window opens that displays the light profile of the selected individual across the entire recording period.
Users can visually inspect the data to determine when the bird likely arrived at and departed from the breeding grounds. These points are typically indicated by a stable pattern of high light intensity (continuous daylight in the Arctic). To specify the range, simply left-click to set the start of the breeding period and right-click to set the end. The function then returns the start and end dates of the selected breeding period for further processing.

*Manual Approach*
```{r, eval = FALSE}
tm.breeding <- set_breeding_period(dir.raw, raw_light, raw_deg, ID, auto = FALSE, gr.Device = "x11")
tm.breeding
```


# 2 - Preprocessing
To create consistent input for analysing the breeding behaviour using a neural network, light and temperature recordings need to be combined into uniform time series with the same step size. The final data frame thus needs to contain four columns with the following names:

1. Date
2. Light
3. Tmin
4. Tmax

Light data provides the reference timeline with the coarsest resolution, while temperature data, often only available in a coarser resolution, needs to be aligned to it. Missing temperature values are interpolated by filling values backward over their original sampling intervals so that they match the resolution of the light data. 

All input variables are then normalised to a range of 0–1 to ensure comparability. To minimise the influence of extreme outliers, realistic value ranges are defined by calculating the 2.5th and 97.5th percentiles of each variable (`Light`, `Tmin`, `Tmax`) across all preprocessed data.

```{r}
segment_days <- 1
breeding_data <- preprocessing(ID, raw_light, raw_deg, tm.breeding, segment_days = segment_days)
```


# 3 - Model (optional)
In this optional section, users can design and train their own deep-learning model for classifying breeding behaviour. Alternatively, a pre-trained model (based on Ruddy Turnstones) can be loaded directly in `Chapter 4 – Application`.
This section allows full flexibility to adjust parameters such as the architecture, training, window length or class weights.
```{r, eval = FALSE}
library(keras3)
library(caret)

# Define window length (number of days)
segment_days <- 1

# Define class weights
weight_brooding <- 2.5
weight_incubation <- 2
weight_random <- 0.8
```

## 3.1 - Creating a training dataset
Before model training, a labelled dataset needs to be created.
Users can gather data from all available geolocator recordings and manually classify representative samples of each behavioural class (brooding, incubation, random). The resulting dataset is then split into training and test subsets for performance evaluation. The following code demonstrates one approach to create and partition such data, but users may freely adapt file structures, proportions, or sampling routines.
```{r, eval = FALSE}
# gather data from all available geolocator records of your species
breeding_data_list <- create_breeding_data_list(dir.raw, segment_days = segment_days)

# Create training data 
preclassified_data <- create_trainingdata(breeding_data_list,
                                          segment_days = segment_days,
                                          dir.raw = dir.raw,
                                          gr.Device = "x11")

# Partition preclassified data into training- (80%) and test-data (20%) 
classes <- sapply(preclassified_data, `[[`, "Class") # extract classes
partition <- caret::createDataPartition(classes, p = 0.8, list = FALSE, times = 1) # 80% of each class goes to training, 20% to test data
training_data <- preclassified_data[partition]
test_data <- preclassified_data[-partition]
```

## 3.2 - Setting up the custom model
If desired, users can define a custom architecture instead of using the pre-trained base model. This step includes preparing the previously created training data, training and evaluation of the model. Below, we provide a template architecture and training procedure that can serve as a starting point for model development.
Users can modify any component — from input features and layer types to kernel sizes, optimisers, or training strategy.

### 3.2.1 - Preparing data for model training
Input data (Light, Tmin, Tmax) are truncated to equal window lengths and combined into 3-channel arrays for model input. You may add or remove variables as you wish.
#### Model arrays
```{r, eval = FALSE}
vars <- c("Light", "Tmin", "Tmax")
window_length <- segment_days * 288

make_array <- function(data, vars, window_length) {
  channels <- lapply(vars, function(var) {
    do.call(rbind, lapply(data, function(x) as.numeric(x[[var]])[seq_len(window_length)]))
  })
  do.call(abind::abind, c(channels, along = 3))
}

x_breeding <- make_array(breeding_data, vars, window_length)
x_train <- make_array(training_data, vars, window_length)
x_test <- make_array(test_data, vars, window_length)
```

### 3.2.2 - Training using 5-fold cross-validation
The example below outlines a standard 5-fold training routine for assessing model robustness. Users can modify the number of folds, metrics, or validation strategies as desired. Each fold’s best model is stored for later comparison.
```{r, eval = FALSE}
model_name <- "custom"
# link directory
dir.model <- file.path(wd, "Model", model_name)

# Create directory inside your data folder for storing the custom model
dir.create(dir.model, recursive = TRUE, showWarnings = FALSE)

# Create k folds
class_levels <- c("brooding", "incubation", "random")
train_classes <- factor(sapply(training_data, `[[`, "Class"), levels = class_levels)
folds <- caret::createFolds(train_classes, k = 5, returnTrain = TRUE)

# K-fold cross-validation
cv_results <- lapply(seq_along(folds), function(i) {
  cat("Processing Fold", i, "of", length(folds), "\n")
  
  # Get indices for this fold
  train_idx <- folds[[i]]
  val_idx <- setdiff(seq_len(length(training_data)), train_idx)

  # Split the data
  x_train_fold <- x_train[train_idx, , , drop = FALSE]
  x_val_fold <- x_train[val_idx, , , drop = FALSE]

  # Prepare labels (one-hot encode)
  y_train_fold <- to_categorical(as.integer(train_classes[train_idx]) - 1,
                                 num_classes = length(class_levels))
  y_val_fold <- to_categorical(as.integer(train_classes[val_idx]) - 1,
                               num_classes = length(class_levels))

  
  # Define the model
  inputs <- layer_input(shape = c(window_length, dim(x_train)[3]))
  outputs <- inputs %>%
    layer_conv_1d(filters = 400, kernel_size = 12, activation = 'relu') %>% 
    layer_average_pooling_1d(pool_size = 12, padding = "same") %>%
    layer_conv_1d(filters = 288, kernel_size = 16, activation = 'relu') %>%
    layer_average_pooling_1d(pool_size = 6, padding = "same") %>%
    layer_flatten() %>%
    layer_dropout(rate = 0.2) %>%
    layer_dense(units = 160, activation = 'relu') %>% #190 or 128
    layer_dense(units = ncol(y_train_fold), activation = 'softmax')
  model <- keras_model(inputs = inputs,
                       outputs = outputs)

  # Compile the model
  model %>% compile(
    loss = 'categorical_crossentropy',
    optimizer = optimizer_adam(
      learning_rate = 0.001,
      beta_1 = 0.9,
      beta_2 = 0.999,
      epsilon = 1e-7
    ),
    metrics = c('categorical_accuracy', 'Precision', 'Recall')
  )

  # Train model
  history <- model %>% keras3::fit(
    x_train_fold, y_train_fold,
    validation_data = list(x_val_fold, y_val_fold),
    epochs = 50,
    batch_size = 32,
    verbose = 0,
    class_weight = list(`0` = weight_brooding,    # brooding
                        `1` = weight_incubation,  # incubation
                        `2` = weight_random),     # random
    callbacks = list(
      callback_early_stopping(monitor = "val_loss", patience = 8, restore_best_weights = TRUE),
      callback_reduce_lr_on_plateau(monitor = "val_loss", factor = 0.5, patience = 4, min_lr = 1e-6),
      callback_model_checkpoint(filepath = file.path(dir.model, paste0("cv_fold_", i, ".keras")), save_best_only = TRUE, monitor = "val_loss")
      )
  )
  
  # Evaluate on validation fold
  val_metrics <- model %>% evaluate(x_val_fold, y_val_fold, verbose = 0)

  val_accuracy  <- as.numeric(val_metrics["categorical_accuracy"])
  val_precision <- as.numeric(val_metrics["Precision"])
  val_recall    <- as.numeric(val_metrics["Recall"])
  val_f1 <- 2 * (val_precision * val_recall) / (val_precision + val_recall)

  cat("Fold", i, "- Val F1:", round(val_f1, 2), "\n")
  keras3::clear_session() # clear between folds

  data.frame(
    fold = i,
    val_accuracy,
    val_precision,
    val_recall,
    val_f1
  )
}) |>
  dplyr::bind_rows()

# Summary of cross-validation results
cat("\n=== Cross-Validation Results ===\n")
print(cv_results)
cat("Mean Validation F1:", round(mean(cv_results$val_f1, na.rm = TRUE), 4), "±", round(sd(cv_results$val_f1, na.rm = TRUE), 4), "\n")
# Precision & Recall = averages over all classes

# Examine CV results for stability and performance
cat("CV Performance Summary:\n")
print(summary(cv_results))

# Check for high variance across folds (indicates instability)
if (sd(cv_results$val_accuracy) > 0.1) {
  cat("Warning: High variance across folds. Consider:\n")
  cat("- More data\n- Different architecture\n- Better regularization\n")
}
```

### 3.2.3 - Final training on full training dataset
Once the best architecture and configuration is identified, the model can be retrained on the entire training dataset to obtain a final model.
```{r, eval = FALSE}
# indicate file path for final model
path_final_model <- file.path(dir.model, "final_model.keras")

# Convert labels to numeric and one-hot encode
classes_training <- sapply(training_data, `[[`, "Class")
y_train <- to_categorical(as.integer(factor(classes_training, levels = class_levels)) - 1,
                          num_classes = length(class_levels))

# Load the best performing model from Cross-Validation
model <- keras3::load_model(file.path(dir.model, paste0("cv_fold_", cv_results$fold[which.max(cv_results$val_f1)], ".keras")))
summary(model)
# Compile the model
model %>% compile(
  loss = 'categorical_crossentropy',
  optimizer = optimizer_adam(
    learning_rate = 0.001,
    beta_1 = 0.9,
    beta_2 = 0.999,
    epsilon = 1e-7),
  metrics = c('categorical_accuracy', 'Precision', 'Recall')
  )

history <- model %>% keras3::fit(
  x_train, y_train,
  epochs = 50,
  batch_size = 64,
  class_weight = list(`0` = weight_brooding,   # brooding
                    `1` = weight_incubation,   # incubation
                    `2` = weight_random),      # random
  callbacks = list(
      callback_early_stopping(monitor = "loss", patience = 3, restore_best_weights = TRUE),
      callback_reduce_lr_on_plateau(monitor = "loss", factor = 0.5, patience = 4, min_lr = 1e-4),
      callback_model_checkpoint(filepath = path_final_model, save_best_only = TRUE, monitor = "loss")    
  )
)
```

## 3.2.4 - Evaluating the model performance on test data
The final model can be validated against the previously created, held-out test set to evaluate the overall model performance.
```{r, eval = FALSE}
# Make predictions on test set
classes_test <- sapply(test_data, `[[`, "Class")
y_test <- to_categorical(as.integer(factor(classes_test, levels = class_levels)) - 1,
                         num_classes = length(class_levels))

test_predictions <- model %>% predict(x_test)
test_metrics <- model %>% evaluate(x_test, y_test)

# Detailed evaluation
pred_classes <- apply(test_predictions, 1, which.max) - 1
true_classes <- apply(y_test, 1, which.max) - 1

# Confusion matrix
cm <- confusionMatrix(
  factor(pred_classes, levels = 0:2, labels = c("brooding", "incubation", "random")),
  factor(true_classes, levels = 0:2, labels = c("brooding", "incubation", "random"))
)
```


# 4 - Application
When there is no need for training or setting up a new model, the pre-trained base model can be directly applied to the preprocessed data. The base model was built using one-day windows of Ruddy Turnstone data. Input data thus also needs to be in the same resolution.
The function `classify_breeding_behaviour` takes the preprocessed dataset and its associated metadata as input and returns a list containing the predicted probabilities for each breeding behaviour state (brooding, incubation, random) across all time windows.
```{r}
classified_breeding <- classify_breeding_behaviour(breeding_data)
classified_breeding
```

Alternatively, if a custom model has been created in Section 3, it can be loaded by specifying its file path via the argument `model = path_final_model`.
```{r, eval = FALSE}
classified_breeding <- classify_breeding_behaviour(breeding_data, model = path_final_model)
classified_breeding
```

## Processing of multiple individuals
Here, we demonstrate how to apply the pre-installed model to breeding data from multiple individuals. Preprocessed daily input periods are stored as a list, classified iteratively, and combined into a single output table containing predicted breeding states and probabilities for all individuals.
```{r, eval = FALSE}
# Define species and directories
Species <- "RuddyTurnstone" 
wd <- "data"
dir.raw <- file.path(wd, "RawData", Species)

# gather data from all available geolocator records of your species
breeding_data_list <- create_breeding_data_list(dir.raw, segment_days = 1)

# remove empty lists
breeding_data_list <- breeding_data_list[lengths(breeding_data_list) > 0]

classified_breeding_df <- lapply(breeding_data_list, classify_breeding_behaviour) |>
  dplyr::bind_rows()

# Save results to CSV
utils::write.csv(
  classified_breeding_df,
  file.path(wd, paste0("classified_breeding_results_", Species, ".csv")),
  row.names = FALSE
)
```
